Abstractions for Type-Level Programming

### 2022-04-08 article.tex
Talk a bit more about type erasure

% \olivier{
% Type erasure is the process by which type arguments of generic types (type constructors) are removed from programs during compilation.
% Scala's erasure affects pattern matching because it prevents certain patterns from being checked at runtime.
% For instance, both `List[String]` and `List[Int]` erase to `List[Object]`, which makes them indistinguishable at runtime.
% This limitation is intrinsic to the JVM and affects Scala's pattern matching in general.
% Match types reduce during type checking (before erasure) and are not affected by this limitation.
% Erasure thus creates an expressivity gap between pattern matching at the term and at the type level.
% }

Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}, where parts of types are preserved and translated to the JVM's type systems, and other parts are erased~\citep{schinz2005compiling}.
For instance, ground classes are compiled directly to JVM classes, but type parameters and type variables are erased and replaced by their bounds.

Erasure directly affects the type tests that can be performed at runtime.
For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it will lead to a compiler warning since the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.

@@

In \SystemFmB (\Cref{subsec:type-binding-extension}), match expressions support two sorts of parametric patterns: they can be either instantiated (\!\!$\match \{ \xs \: \List~\Int \}$, where $\Int \∈ \A$), or use a binding pattern variable (\!\!$\match \[\X] \{ \xs \: \List~\X \}$, where $\X$ is a pattern variable).
In this sense, \SystemFmB is \emph{more expressive} than Scala, where instantiated patterns are not available at the term level due to type erasure.
% TODO: say that our approach would not work in a system with full type erasure

>>>>

Scala's primary platform is the Java virtual machine (JVM).
On the JVM, Scala is compiled using \emph{partial erasures}~\citep{schinz2005compiling}, a process which preserves some types by translating them to the JVM's type systems, and removes other types by erasing them to |Object|, the JVM's top type.
Scala's type erasure directly affects pattern matching because it prevents certain patterns from being checked at runtime.
For instance, both |List[String]| and |List[Int]| erase to |List[Object]|, which makes them indistinguishable from the JVM's perspective.
This limitation is intrinsic to the JVM and affects Scala's pattern matching in general.
Match types reduce during type checking (before erasure) and are thus not affected by this limitation.
Erasure thus creates an expressivity gap between pattern matching at the term and at the type level.

In practice, the Scala compiler copes with this limitation by emitting warnings for unchecked patterns.
For example, while |case xs: List[Int] =>| is a syntactically valid pattern, it compiles with a warning the |Int| type parameter is eliminated by erasure and cannot be checked at runtime.

@@

In \SystemFmB (\Cref{subsec:type-binding-extension}), match expressions support two sorts of parametric patterns: they can be either instantiated (\!\!$\match \{ \xs \: \List~\Int \}$, where $\Int \∈ \A$), or use a binding pattern variable (\!\!$\match \[\X] \{ \xs \: \List~\X \}$, where $\X$ is a pattern variable).
In this sense, \SystemFmB is \emph{more expressive} than Scala, where instantiated patterns are not available at the term level due to type erasure.

In a language with full type erasure, match types would have no term-level counterparts and the \TMatch rule would thus be pointless.

### 2022-04-08 symposium.tex
Address Martin's comments on abmiguities
Add Richard's comments
Talk a bit more about type erasure

% TODO: consider inline functions as a 3rd alternative

@@

Stephanie Weirich presented on several occasions a regular expression library as an example of a dependently-typed program written in Haskell \citep{weirich2014examples}.
The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
To make that possible, her implementation makes heavy use of Haskell's advanced type system features and requires no less than 20 language extensions to compile.
In contrast, our implementation does not use metaprogramming and compiles with vanilla Scala.
An important conceptual difference is that our approach only focuses on type checking (we rely on Java's regular expressions at run-time), whereas her library also contains a complete regular expression engine.

The typed-regex~\citep{akshay2021typed} Typescript library provides a type-safe API for extracting named capturing groups of regular expressions.
The implementation uses conditional types~\citep{microsoft2020typescript}, Typescript's type-level ternary operator, to statically analyze regular expressions.
Their approach, like ours, identifies optional capturing groups and marks them as such (using Typescript's optional properties).
At the time of writing, the library is still in the early stages of development and only supports a subset of the regular expression syntax, and notably lacks support for nested optional capturing group detection.

>>>>

@@

Stephanie Weirich presented on several occasions a regular expression library as an example of a dependently-typed program written in Haskell \citep{weirich2014examples}.
The library provides safety guarantees similar to ours, but from the slightly different angle of named-capturing groups.
% TODO: Rework that sentence
An important conceptual difference is that our approach only focuses on type checking (we rely on Java's regular expressions at run-time), whereas her library also contains a complete regular expression engine.

The typed-regex~\citep{akshay2021typed} Typescript library provides a type-safe API for extracting named capturing groups of regular expressions.
The implementation uses conditional types \citep{microsoft2020typescript}, Typescript's type-level ternary operator, to statically analyze regular expressions.
Their approach, like ours, identifies optional capturing groups and marks them as such (using Typescript's optional properties).
At the time of writing, the library is still in the early stages of development and only supports a subset of the regular expression syntax, and notably lacks support for nested optional capturing group detection.
